---
title: "R Day 2"
author: "Michael Fedell"
date: "9/18/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


To access a single function or resource from a package not yet loaded, use
<package_name>::<resource_name>

here::here() is useful for accessing current pwd - e.g. so long as repo is the same for all users, does not matter where repo exists in filesystem running project
```{r}
here::here('data', 'gapminder5.csv')

gapminder <- read.csv(here::here('data', 'gapminder5.csv'))
str(gapminder)
```

## Converting factors to characters

```{r}
gapminder$country <- as.character(gapminder$country)
gapminder$continent <- as.character(gapminder$continent)
str(gapminder)
```

## Reporting average life expectancy per country

To find the average life expectancy for a country, can subset the data like:
```{r}
gapminder$lifeExp[gapminder$country == 'Albania']

gapminder$lifeExp[gapminder$country == 'Afghanistan']
```

But to view all, set up a loop for each country in the data.
```{r}
for (country in gapminder$country) {
  (gapminder$lifeExp[gapminder$country == country])
}
```

*Note!:* Avoid for loops in R when possible, they are expensive and slow!

Let's recover GDP for each country
```{r}
# create a vector of values that you want to repeat the function for
obs <- 1:nrow(gapminder)

# initialize the for loop with `for (i in vector)`
for (i in obs) { # the function to repeat is enclosed in braces
  gapminder[i, 'gdp'] <- gapminder[i, 'pop'] * gapminder[i, 'gdpPercap']
}
```

And do some normalization as well!
```{r}
for (i in obs) { # the function to repeat is enclosed in braces
  gapminder[i, 'log_gdpPercap'] <- log(gapminder[i, 'gdpPercap'])
  gapminder[i, 'log_pop'] <- log(gapminder[i, 'pop'])
}
str(gapminder)
```

```{r}
years <- unique(gapminder$year)

for (year in years) {
  mean_le <- mean(gapminder$lifeExp[gapminder$year == year],
                  na.rm=T)
  print(paste0(year, ': ', mean_le))
}
```

Find continent with the greatest mean life expectancy
```{r}
continents <- unique(gapminder$continent)

for (cont in continents) {
  mean.le <- mean(gapminder$lifeExp[gapminder$continent == cont], na.rm=T)
  print(paste0(cont, ': ', mean.le))
}
```

```{r}
for (cont in continents) {
  print(cont)
  for (year in years) {
    le.dev <- sd(gapminder$lifeExp[gapminder$continent == cont & gapminder$year == year])
    print(paste0('life exp SD in ', year, ': ', le.dev))
  }
}
```

```{r}
vars <- gapminder[, c("lifeExp", "pop", "gdpPercap")]
apply(vars, 2, mean)
```

```{r}
lapply(gapminder, mean)
```

```{r}
sapply(gapminder, mean)
```

```{r}
sapply(years, function(x) mean(gapminder$lifeExp[gapminder$year == x]))
```

```{r}
year <- 1987

while (year <= 2002) {
  sd.le <- sd(gapminder$lifeExp[gapminder$year == year])
  print(paste0(year, ': ', sd.le))
  year <- year + 5
}
```

```{r}
for (year in unique(gapminder$year)) {
  if (year >= 1987) {
    print(paste0(year, ': ', mean(gapminder$pop))) 
  } else {
    print(paste0(year, ' out of range'))
  }
}
```

```{r}
get_values <- function(df, col='continent') {
  vals <- unique(df[[col]])
  print(paste0(col, ': ', vals))
  return(vals)
}
```

```{r}
report.stats <- function(df, cont) {
  le.mean <- mean(gapminder$lifeExp[gapminder$continent == cont])
  le.med <- median(gapminder$lifeExp[gapminder$continent == cont])
  le.min <- min(gapminder$lifeExp[gapminder$continent == cont])
  le.max <- max(gapminder$lifeExp[gapminder$continent == cont])
  cat('Life Expectancy for: ', cont,
      '\n\tMean: ', le.mean,
      '\n\tMedian: ', le.med,
      '\n\tMin: ', le.min,
      '\n\tMax: ', le.max)
}
```
